Plume-SPH 1.0: A three-dimensional, dusty-gas volcanic plume model based on smoothed particle hydrodynamics

### 2017-05-06 Manuscripts_GMD.tex
update Dr. Patra's modification

%>>>>>>> 0d96d9682fb484b8b159e8d09a0eb89723d3a0f8
% \Author[affil]{given_name}{surname}
\Author[1]{Zhixuan}{Cao}
\Author[1]{Abani}{Patra}
\Author[2]{Marcus}{Bursik}
\Author[3]{E. Bruce}{Pitman}
\Author[4]{Matthew}{Jones}
\affil[1]{Department of Mechanical and Aerospace Engineering, University at Buffalo, SUNY, New York, United State}
\affil[2]{Department of Geology, University at Buffalo, SUNY, New York, United State}
\affil[3]{Department of Material Design and Innovation, University at Buffalo, SUNY, New York, United State}
\affil[4]{Center for Computational Research,
University at Buffalo, SUNY, New York, United State}
%% The [] brackets identify the author with the corresponding affiliation. 1, 2, 3, etc. should be inserted.
\runningtitle{Plume-SPH 1.0}
\runningauthor{Z. Cao et al}
%<<<<<<< HEAD
\correspondence{Abani Patra (abani@buffalo.edu)}
%=======
%\correspondence{Abani Patra (abani.patra@gmail.com)}
%>>>>>>> 0d96d9682fb484b8b159e8d09a0eb89723d3a0f8
\received{}
\pubdiscuss{} %% only important for two-stage journals
\revised{}
\accepted{}
\published{}
%% These dates will be inserted by Copernicus Publications during the typesetting process.
\firstpage{1}
\maketitle
\begin{abstract}
VATDs (Volcanic ash transport and dispersal models) used for volcanic ash forecasts sometimes take outputs of eruption plume models as source terms. The accuracy of source terms is crucial for forecasts from VATDs. All existing 3D (three dimensional) plume models described in the literature use mesh based methods. SPH (smoothed particle hydrodynamics), as a mesh free method, has several advantages over mesh based methods in modeling of multiphase free boundary flows. As an initial effort to exploit the feasibility and advantages of SPH in volcanic plume modeling, we adopt a relatively simple model, a 3D dusty-gas dynamic model, targeted at capturing the salient features of a volcanic plume. In this model, erupted material is assumed to be well-mixed, and represented by one phase while air is another phase. We also assume immediately dynamics and thermodynamics equilibrium between air and erupted material and that the effect of wind is not significant.

Several newly developed techniques in SPH are adopted and extended to address numerical challenges in simulating multiphase compressible turbulent flow with classical SPH method. The $SPH-\varepsilon$ turbulence model is adopted to capture mixing at unresolved scales. Heat exchange due to turbulence is then calculated with a Reynolds analogy. A corrected formalism of SPH is used to handle tensile instability and deficiency of particle distribution near the boundaries. We also propose feasible ways to impose velocity inlet and pressure outlet boundary conditions, both of which are scarce in traditional implementations of SPH. The core solver of our model is parallelized with MPI (message passing interface) obtaining good weak-scalability and strong-scalability. The model is first verified by comparing velocity and concentration distribution along axis and in cross section with experimental results of JPUE (jet or plume that is ejected from a nozzle into a uniform environment). Next, the top height of the Pinatubo eruption of 15 June 1991, as simulated by our model, is consistent with both observations and existing 3D plume models. Profiles of several integrated variables are compared with those calculated in existing 3D plume models, and further verify our model. Analysis of the plume evolution process illustrates that this model is able to reproduce the basic physics of plume development. The comparison also implies that the turbulence model plays a significant role in 3D volcanic plume modeling.
\end{abstract}

@@

\subsection{Existing plume models}
Several 1D (one dimensional) volcanic plume models have been developed in the past few decades, ranging from the most basic 1D model \citep{woods1988fluid} which only accounts for mass conservation to more recently developed 1D models  \citep{bursik2001effect, mastin2007user, degruyter2012improving, woodhouse2013interaction, devenish2013using, de2015plume, folch2016fplume, pouget2016sensitivity} which tend to account for more comprehensive physics effects. 
For example, FPLUME-1.0 \citep{folch2016fplume} accounts for wind effect, entrainment of moisture, water phase change, particle fallout and re-entrainment and even wet aggregation of ash. However, in these 1D models, the entrainment of air is evaluated based on two coefficients: entrainment coefficient for the vertical plume and the entrainment coefficient that describes the effect of wind. Different 1D models adopt different entrainment coefficients based on specific formulation or calibration against well-documented case studies. The feedback from plume to atmosphere is also ignored in 1D models. Even though determination of essential parameters such as the entrainment is not based on first principles, such simple models nevertheless allow us to investigate the importance of physical mechanisms in a volcanic plume. In addition, these simplified models require little computational resource and can run on standard personal computers or on web sites in very short time. As a result, 1D software for volcanic plume development \citep[such as][]{267, 1194, 3541}, combined with VATDs \citep[such as][]{114, draxler2015hysplit} are widely used in research and practice. While these 1D models can generate well-matched results with 3D (three dimensional) models for weak plumes, much greater variability is observed for strong plume scenarios, especially for local variables \citep{costa2016results}. In addition, the need for greater skill in hazards forecasts especially where the plume model is used to generate source conditions for complex VATD models requires 2D (two dimensional) and 3D models.

The development of 2D and 3D, time-dependent, and multiphase numerical models for volcanic plumes has provided new explanations for many features of explosive volcanism. One of the earliest of these is the 3D model PDAC (Pyroclastic Dispersion Analysis Code) \citep{neri2003multiparticle}  which is a non-equilibrium, multiphase, 3D compressible flow model. Conservation equations for each phase are solved separately with the finite volume method. A parallel computing version of PDAC was developed lately \citep{ongaro2007parallel}. Advanced numerical techniques, such as second order scheme and semi-explicit time stepping, was also adopted afterwards to improve the accuracy of PDAC \citep{carcano2013semi}.

@@

Besides adding to their special strengths (ATHAM has been adding more and more microphysics, PDAC was extended to consider more phases), these models are learning from each other. PDAC development has begun to include the effect of microphysics into the model while ATHAM development has extended its ability to modeling pyroclastic flow. Both are using finer and finer resolution.

@@

\subsection{Features of SPH}
To the best of our knowledge, all of the existing 3D plume models are mesh based Eulerian methods, and there are no 3D plume models based on mesh free Lagrangian methods. Lagrangian methods have several features that are suitable for volcanic plume simulation that we outline below. Among such Lagrangian methods, smoothed particle hydrodynamics \citep{gingold1977smoothed,lucy1977numerical} based simulations have shown good agreement with experiments for many applications in fluid dynamics. And it currently is, by far, the most widely used mesh-free scheme. Our implementations of SPH in volcanology follows on earlier efforts   \citep{bursik2003smoothed,herault2010sph,haddad2016smoothed}.
Specifically, we choose SPH as the numerical method for volcanic plume simulation to enable:
\begin{itemize}
\item Better investigation of mixing phenomena;
\item accurate modeling of the development of ZFE (Zone of Flow Establishment), ZEF (Zone of Established Flow) investigation and relation to column collapse and the questions relating to the development of entrainment; %and proper characterization through numerical experimentation.
\item  easy inclusion of particles of different sizes (phases) and investigation of detailed mechanics of sedimentation and drag force interaction in lower plume.
\end{itemize}
These are enabled by the following features of SPH:
\begin{itemize}
\item For mesh based method, either interface tracking (Lagrangian) \citep{harlow1965numerical, wrobel1991computational, cheng1995simplified} or interface capturing (Eulerian) \citep{hirt1981volume, youngs1982time, gerlach2006comparison, gopala2008volume} methods are used to reconstruct the flow interface of free boundary flow. High computational cost, a tendency to form numerical instabilities and the inability to track complex topological changes are the significant drawbacks of tracking techniques \citep{hirt1981volume, unverdi1992front, anderson1998diffuse}. For interface capturing (Eulerian) method, the surrendering of surface detail before the phase transport calculation means that interface reconstruction is required between time steps to recover the interface information, which needs additional numerical effort \citep{hirt1981volume, youngs1982time}. Since SPH is able to adaptively adjust the discretization and automatically construct the interface, SPH does not require additional numerical effort for interface construction and therefore is more suitable for volcanic plume simulation.
\item Advection term in the Navier-Stokes equations does not appear explicitly in discretized formula of SPH (as illustrated in Eq. (\ref{eq:gov-nc-rho}) to Eq. (\ref{eq:gov-nc-e})).
\item It is easy to include various physics effect (like self gravity, radiative cooling and chemical reaction) into the model. It does not require a major overhaul and re-tooling every time new physics is introduced \citep{monaghan1995sph}. This implies that accounting for more physics is easier for SPH model.
\item With more than one material, each described by its own set of particles, interface problems between phases are often trivial for SPH but difficult for mesh based schemes. So multiphase flow can be easily handled by SPH. Adding of new phases to the model also does not require a major overhaul and re-tooling. As will be shown in later paragraphs, adding of new phases only leads to adding of several lines into the source code for new phases and additional interaction terms between existing phases and newly added phases.
\end{itemize}

As discussed in the previous paragraph, existing 3D plume models focus on one or several specific aspects of plume and have been extended to be more comprehensive by accounting for more physics or more phases. Easy extensibility and capability of handling multiples phase flow without additional numerical effort greatly facilitate future extension of SPH models. As volcanic plumes are in nature multiphase and without pre-defined boundary in the atmosphere, SPH is a suitable numerical method for plume modeling. The core physics, such as entrainment of air and thermal expansion, are essential for all plume modeling while some other physics, such as water condensation and aggregation etc., are important in specific scenarios. As an initial effort on exploiting advantages of SPH in volcanic plume modeling, we focus on capturing basic features in plume development using a robust numerically and comuptationaly efficient framework with support for scalable parallel computing. %So our current model is far away from comprehensive.

@@

\begin{itemize}
\item We develop methodology to impose pressure boundary conditions by adding extra layers of static ghost particles. Additional constraints on the time step is used to avoid the growth of numerical fluctuations near the pressure boundary. We impose a velocity inlet boundary condition by placing several layers of ghost particles moving with eruption velocity. 
\item Turbulence model is crucial for reproducing the entrainment of air. There are several turbulence models proposed for SPH method \citep{issa2005numerical,violeau2007numerical}. We adopt a LANS (Lagrangian Averaged Navier-Stokes) turbulence model \citep{monaghan2011turbulence}, which was originally proposed for incompressible flow, and is extended here for compressible flow accounting for turbulent heat exchange. 
\item Corrected formulation of SPH \citep{chen1999improvement} is adapted to bypass the well-known tensile instability issues of classical SPH.
\item Simulation of volcanic plumes with acceptable accuracy requires fine resolution (very high particle counts) that cannot be accomplished without parallel computing. The core solver of our model is parallelized by MPI (message passing interface standard) parallelism. In addition, a dynamic load balancing strategy is also developed. 
\item Imposition of some types of boundary conditions (such as eruption boundary condition) requires dynamically adding and removing of particles during simulation. To address this issue, we adopt an efficient data management scheme based on time dependent SFC (space filling curve) and hash table. The computational cost is further reduced by adjusting simulation domain adaptively.
\end{itemize}

@@

In mesh based methods, governing equations in Eulerian form, Eq. (\ref{eq:gov-cs-rho}) to Eq. (\ref{eq:gov-cs-e}), are directly discretized. For SPH, governing equations in Lagrange form is needed. By deducting kinetic energy from energy equation, subtracting mass conservation from momentum equation, combining transient term and advection term into material derivative term, the governing equations are put into the final form, in which advection term does not show up explicitly.
\begin{align}
\dfrac{D \rho}{D t} + \rho \nabla \cdot \textbf{v} = 0 \label{eq:gov-nc-rho}\\
\dfrac{D \rho \xi}{D t} + \rho \xi \nabla \cdot \textbf{v} = 0 \label{eq:gov-nc-ks}\\
\dfrac{D \textbf{v}}{D t} + \dfrac{\nabla P}{\rho} =\textbf{g} \label{eq:gov-nc-v}\\
\dfrac{D e}{D t} + \dfrac{P \nabla \cdot \textbf{v}}{\rho} = 0 \label{eq:gov-nc-e}
\end{align}
%Governing equations in Lagrangian form, Eq. (\ref{eq:gov-nc-rho}) to Eq.(\ref{eq:gov-nc-e}), and Eulerian form, Eq. (\ref{eq:gov-cs-rho}) to Eq. (\ref{eq:gov-cs-e}), are actually equivalent, as deriving the Lagrangian form from the Eulerian is purely based on mathematics without any additional physical assumptions.

@@

\section{SPH method} \label{sec:SPH_method}
SPH is a mesh-free Lagragian method. In SPH, the domain (control mass) is discretized by a set of particles or discretization points and the position of each particle (discretization point) is updated at every time step based on the motion computed. Approximation of all field variables (velocity, density and pressure, ect.) is obtained by interpolation based on discretization points. The physical laws (such as conservation laws of mass, momentum and energy) written in the form of PDEs (partial differential equations)  or ODEs (ordinary differential equations) need to be transformed into the Lagrangian particle formalism of SPH. Using a kernel function that provides the weighted estimation of the field variables at any point, the integral equations are estimated as sums over particles in a compact subdomain defined by the support of the kernel function defined with respect to the discretization points. Thus, field variables associated to the particle are updated based on its neighbors. Each kernel function has a compact support determined by smoothing length of each particle. There are several review papers by \citet{monaghan1992smoothed, monaghan2005smoothed, rosswog2009astrophysical, price2012smoothed, monaghan2012smoothed}, giving a pretty comprehensive view over SPH, we only refer here to the representation of the constitutive equations in SPH and put more focus on specific numerical techniques for plume modeling.

\subsection{Fundamental principles}
There are several procedures for discretizing governing equations (PDEs or ODEs) with SPH. There is slight difference between discretized formalism obtained by different routes. We present here one of them following \citet{monaghan1992smoothed, monaghan2005smoothed, monaghan2012smoothed}. The starting point of approximating a function with SPH is the translation property of the Dirac function.
\begin{equation}
A\left(\textbf{x}\right)=\int_{-\infty}^{\infty} A\left(\textbf{x} \prime\right) \delta \left(\textbf{x} \prime - \textbf{x}\right) d\textbf{x} \prime
\label{eq:Dirac-translation}
\end{equation}
The Dirac function in Eq. (\ref{eq:Dirac-translation}) can be approximated by a weighting function $w\left(\textbf{x}-\textbf{x}\prime, h\right)$ (or $w\left(\textbf{x}\prime-\textbf{x}, h\right)$)) which tends to a Dirac function when the smoothing length $h \rightarrow 0$ :
\begin{equation}
\lim _{h \rightarrow 0} w\left(\textbf{x} \prime-\textbf{x}, h\right) =  \delta \left(\textbf{x} \prime - \textbf{x}\right)
\label{eq:SPH_kernel_delta}
\end{equation}
An arbitrary function $A\left(\textbf{x}\right)$ can then be approximated by:
\begin{equation}
A\left(\textbf{x}\right) \approx <A\left(\textbf{x}\right)> = \int_{\Omega} A\left(\textbf{x} \prime\right) w\left(\textbf{x}-\textbf{x}\prime, h\right) d\textbf{x}\prime + O\left(h^2\right)
\label{eq:SPH-fundamental-principle}
\end{equation}
Where $h$ is the smoothing length, determining the interaction distance. As the weighting function is symmetric (Eq. (\ref{eq:SPH-kernel-symmetric})) and satisfies the normalization condition (Eq. (\ref{eq:SPH-kernel-normalization-prop})), odd error terms in Eq. (\ref{eq:SPH-fundamental-principle}) vanishe leading to a second order approximation. However, in practice, second order of accuracy can not be achieved because there is no guarantee on the symmetry of particle distribution in real simulation \citep{price2012smoothed}.
Recall that $d\textbf{x}\prime = \dfrac{dm \prime}{\rho \prime}$, the integration equation, Eq. (\ref{eq:SPH-fundamental-principle}), can be approximated by summation and lead to an approximation of the function $A$:
\begin{equation}
<A\left(\textbf{x}\right)> \approx \sum_b m_b \dfrac{A_b}{\rho_b} w\left(\textbf{x}-\textbf{x}_b, h\right)
\label{eq:SPH-approximation-sum}
\end{equation}
where the summation is over all the particles within the region of compact support of the weighting function. 
Gradient terms may be straightforwardly calculated by taking the derivative of Eq. (\ref{eq:SPH-approximation-sum}), giving
\begin{equation}
\begin{split}
<\nabla A\left(\textbf{x}\right)> & = \dfrac{\partial }{\partial \textbf{x}} \int_{\Omega} A\left(\textbf{x} \prime\right) w\left(\textbf{x}-\textbf{x}\prime, h\right) d\textbf{x}\prime + O\left(h^2\right) \\
& \approx \sum_b m_b \dfrac{A_b}{\rho_b} \nabla w\left(\textbf{x} - \textbf{x}_b, h\right)
\end{split} 
\label{eq:SPH-scalar-function-gradient}
\end{equation}
For vector quantities the expressions are similar, simply replacing $A$ with $\textbf{A}$ in Eq. (\ref{eq:SPH-approximation-sum}) and Eq. (\ref{eq:SPH-scalar-function-gradient}), giving
\begin{align}
<\textbf{A}\left(\textbf{x}\right)> \approx \sum_b m_b \dfrac{\textbf{A}_b}{\rho_b} w\left(\textbf{x}-\textbf{x}_b, h\right) \\
<\nabla \cdot \textbf{A}\left(\textbf{x}\right)> \approx \sum_b m_b \dfrac{\textbf{A}_b}{\rho_b} \cdot \nabla w\left(\textbf{x} - \textbf{x}_b, h\right) \\
<\nabla \times \textbf{A}\left(\textbf{x}\right)> \approx \sum_b m_b \dfrac{\textbf{A}_b}{\rho_b} \times \nabla w\left(\textbf{x} - \textbf{x}_b, h\right) \\
<\nabla^j \textbf{A}^i\left(\textbf{x}\right)> \approx \sum_b m_b \dfrac{\textbf{A}_b^i}{\rho_b} \nabla^j w\left(\textbf{x} - \textbf{x}_b, h\right) 
\label{eq:SPH-vecctor-function}
\end{align}
%

>>>>

%>>>>>>> 0d96d9682fb484b8b159e8d09a0eb89723d3a0f8
% \Author[affil]{given_name}{surname}
\Author[1]{Zhixuan}{Cao}
\Author[1]{Abani}{Patra}
\Author[2]{Marcus}{Bursik}
\Author[3]{E. Bruce}{Pitman}
\Author[4]{Matthew}{Jones}
\affil[1]{Department of Mechanical and Aerospace Engineering, University at Buffalo, SUNY, New York, United State}
\affil[2]{Department of Geology, University at Buffalo, SUNY, New York, United State}
\affil[3]{Department of Material Design and Innovation, University at Buffalo, SUNY, New York, United State}
\affil[4]{Center for Computational Research,
University at Buffalo, SUNY, New York, United State}
%% The [] brackets identify the author with the corresponding affiliation. 1, 2, 3, etc. should be inserted.
\runningtitle{Plume-SPH 1.0}
\runningauthor{Z. Cao et al}
%<<<<<<< HEAD
\correspondence{Abani Patra (abani@buffalo.edu)}
%=======
%\correspondence{Abani Patra (abani.patra@gmail.com)}
%>>>>>>> 0d96d9682fb484b8b159e8d09a0eb89723d3a0f8
\received{}
\pubdiscuss{} %% only important for two-stage journals
\revised{}
\accepted{}
\published{}
%% These dates will be inserted by Copernicus Publications during the typesetting process.
\firstpage{1}
\maketitle
\begin{abstract}
VATDs (Volcanic ash transport and dispersal models) used for volcanic ash forecasts often take outputs of eruption plume models as source terms. The accuracy of these terms is crucial for forecasts from VATDs. All existing 3D (three dimensional) plume models described in the literature use mesh based methods. SPH (smoothed particle hydrodynamics), as a mesh free method, has several advantages over mesh based methods in modeling of multiphase free boundary flows. As an initial effort to exploit the feasibility and advantages of SPH in volcanic plume modeling, we adopt a relatively simple model, a 3D dusty-gas dynamic model, targeted at capturing the salient features of a volcanic plume. In this model, erupted material is assumed to be well-mixed, and represented by one phase while air is another phase. We also assume dynamic and thermodynamical equilibrium between air and erupted material and that the effect of wind is not significant.

Several newly developed techniques in SPH are adapted and extended to address numerical challenges in simulating multiphase compressible turbulent flow with classical SPH method. The $SPH-\varepsilon$ turbulence model is to capture mixing at unresolved scales. Heat exchange due to turbulence is then calculated by Reynolds analogy. A corrected SPH is used to handle tensile instability and deficiency of particle distribution near the boundaries. We also propose to impose velocity inlet and pressure outlet boundary conditions, both of which are scarce in traditional implementations of SPH. The core solver of our model is parallelized with MPI (message passing interface) obtaining good weak and strong scalability. The model is verified by comparing velocity and concentration distribution along axis and in cross section with experimental results of JPUE (jet or plume that is ejected from a nozzle into a uniform environment). Next, the top height of the Pinatubo eruption of 15 June 1991, as simulated by our model, is consistent with both observations and existing 3D plume models. Profiles of several integrated variables are compared with those calculated in existing 3D plume models, and further verify our model. Analysis of the plume evolution process illustrates that this model is able to reproduce the basic physics of plume development. The comparison also implies that the turbulence model plays a significant role in 3D volcanic plume modeling.
\end{abstract}

@@

\subsection{Existing plume models}
Several 1D (one dimensional) volcanic plume models have been developed in the past few decades, ranging from the most basic 1D model \citep{woods1988fluid} which only accounts for mass conservation to more recently developed 1D models  \citep{bursik2001effect, mastin2007user, degruyter2012improving, woodhouse2013interaction, devenish2013using, de2015plume, folch2016fplume, pouget2016sensitivity} which tend to account for more comprehensive physics effects. 
For example, FPLUME-1.0 \citep{folch2016fplume} accounts for wind effect, entrainment of moisture, water phase change, particle fallout and re-entrainment and even wet aggregation of ash. However, in these 1D models, the entrainment of air is evaluated based on two coefficients: entrainment coefficient for the vertical plume and the entrainment coefficient that describes the effect of wind. Different 1D models adopt different entrainment coefficients based on specific formulation or calibration against well-documented case studies. The feedback from plume to atmosphere is usually ignored in 1D models. Even though determination of essential parameters such as the entrainment is not based on first principles, such simple models nevertheless allow us to investigate the importance of physical mechanisms in a volcanic plume. In addition, these simplified models require little computational resource and can run on standard personal computers or on web sites in very short time. As a result, 1D software for volcanic plume development \citep[such as][]{267, 1194, 3541}, combined with VATDs \citep[such as][]{114, draxler2015hysplit} are widely used in research and practice. While these 1D models can generate well-matched results with 3D (three dimensional) models for weak plumes, much greater variability is observed for strong plume scenarios, especially for local variables \citep{costa2016results}. In addition, there is need for greater skill in hazards forecasts especially where the plume model is used to generate source conditions for complex 2D (two dimensional) and 3D VATD models.

The development of 2D and 3D, time-dependent, and multiphase numerical models for volcanic plumes has provided new explanations for many features of explosive volcanism. One of the earliest of these is the 3D model PDAC (Pyroclastic Dispersion Analysis Code) \citep{neri2003multiparticle}  which is a non-equilibrium, multiphase, 3D compressible flow model. Conservation equations for each phase are solved separately with the finite volume method. A parallel computing version of PDAC was also developed \citep{ongaro2007parallel}. Advanced numerical techniques, such as a second order scheme and semi-explicit time stepping, was also adopted afterwards to improve the accuracy of PDAC \citep{carcano2013semi}.

@@

Besides adding to their special strengths (ATHAM has been adding more and more microphysics, PDAC was extended to consider more phases), these models are also adding core strengths. PDAC development has begun to include the effect of microphysics into the model while ATHAM development has extended its ability to modeling pyroclastic flow. Both are using finer and finer resolution.

@@

\subsection{Features of SPH}
To the best of our knowledge, all of the existing 3D plume models are mesh based Eulerian methods, and there are no 3D plume models based on mesh free Lagrangian methods. Lagrangian methods have several features that are suitable for volcanic plume simulation that we outline below. Among such Lagrangian methods, smoothed particle hydrodynamics \citep{gingold1977smoothed,lucy1977numerical} based simulations have shown good agreement with experiments for many applications in fluid dynamics. And it currently is, by far, the most widely used mesh-free scheme. Our implementations of SPH in volcanology follows on earlier efforts   \citep{bursik2003smoothed,herault2010sph,haddad2016smoothed}.
Specifically, we choose SPH as the numerical method for volcanic plume simulation to enable:
\begin{itemize}
\item better investigation of mixing phenomena;
\item accurate modeling of the development of ZFE (Zone of Flow Establishment), ZEF (Zone of Established Flow) investigation and relation to column collapse and the questions relating to the development of entrainment; %and proper characterization through numerical experimentation.
\item  easy inclusion of particles of different sizes (phases) and investigation of detailed mechanics of sedimentation and drag force interaction in lower plume.
\end{itemize}
These are enabled by the following features of SPH:
\begin{itemize}
\item For mesh based method, either interface tracking (Lagrangian) \citep{harlow1965numerical, wrobel1991computational, cheng1995simplified} or interface capturing (Eulerian) \citep{hirt1981volume, youngs1982time, gerlach2006comparison, gopala2008volume} methods are used to reconstruct the flow interface of free boundary flow. High computational cost, a tendency to form numerical instabilities and the inability to track complex topological changes are the significant drawbacks of tracking techniques \citep{hirt1981volume, unverdi1992front, anderson1998diffuse}. For interface capturing (Eulerian) method, the surrendering of surface detail before the phase transport calculation means that interface reconstruction is required between time steps to recover the interface information, which needs additional numerical effort \citep{hirt1981volume, youngs1982time}. Since SPH is able to adaptively adjust the discretization and automatically construct the interface, SPH does not require additional numerical effort for interface construction and therefore is more suitable for volcanic plume simulation.
\item Advection term in the Navier-Stokes equations does not appear explicitly in discretized formula of SPH (as illustrated in Eq. (\ref{eq:gov-nc-rho}) to Eq. (\ref{eq:gov-nc-e})).
\item It is easy to include various physics effects (like self gravity, radiative cooling and chemical reaction) into the model. It does not require a major overhaul and re-tooling every time new physics is introduced \citep{monaghan1995sph}. This implies that accounting for more physics is easier for SPH model.
\item With more than one material, each described by its own set of particles, interface problems between phases are often trivial for SPH but difficult for mesh based schemes. So multiphase flow can be easily handled by SPH. Adding of new phases to the model also does not require a major overhaul and re-tooling. As will be shown in later paragraphs, adding of new phases only leads to adding of several lines into the source code for new phases and additional interaction terms between existing phases and newly added phases.
\end{itemize}

As discussed in the previous paragraph, existing 3D plume models focus on one or several specific aspects of plume and have been extended to be more comprehensive by accounting for more physics or more phases. Easy extensibility and capability of handling multiples phase flow without additional numerical effort greatly facilitate future extension of SPH models. As volcanic plumes are in nature multiphase and without pre-defined boundary in the atmosphere, SPH is a suitable numerical method for plume modeling. The core physics, such as entrainment of air and thermal expansion, are essential for all plume modeling while some other physics, such as water condensation and aggregation etc., are important in specific scenarios. As an initial effort on exploiting advantages of SPH in volcanic plume modeling, we focus on capturing basic features in plume development using a robust numerically and computationally efficient framework with support for scalable parallel computing. %So our current model is far away from comprehensive.

@@

\begin{itemize}
\item We develop methodology to impose pressure boundary conditions by adding extra layers of static ghost particles. Additional constraints on the time step is used to avoid the growth of numerical fluctuations near the pressure boundary. We impose a velocity inlet boundary condition by placing several layers of ghost particles moving with eruption velocity. 
\item Turbulence model is crucial for reproducing the entrainment of air. There are several turbulence models proposed for SPH method \citep{issa2005numerical,violeau2007numerical}. We adopt a LANS (Lagrangian Averaged Navier-Stokes) turbulence model \citep{monaghan2011turbulence}, which was originally proposed for incompressible flow, and is extended here for compressible flow accounting for turbulent heat exchange. 
\item Corrected formulation of SPH \citep{chen1999improvement} is adapted to bypass the well-known tensile instability issues of classical SPH.
\item Simulation of volcanic plumes with acceptable accuracy requires fine resolution (very high particle counts) that cannot be accomplished without parallel computing. The core solver of our model is parallelized by MPI (message passing interface standard) parallelism. In addition, a dynamic load balancing strategy is also developed. 
\item Imposition of some types of boundary conditions (such as eruption boundary condition) requires dynamically adding and removing of particles during simulation. To address this issue, we adopt an efficient data management scheme based on time dependent SFC (space filling curve) induced indexing and hash table. The computational cost is further reduced by adjusting simulation domain adaptively.
\end{itemize}

@@

In mesh based methods, governing equations in Eulerian form, Eq. (\ref{eq:gov-cs-rho}) to Eq. (\ref{eq:gov-cs-e}), are directly discretized. For SPH, governing equations in Lagrange form are needed. By deducting kinetic energy from energy equation, subtracting mass conservation from momentum equation, combining transient term and advection term into material derivative term, the governing equations are put into the final form, in which advection term does not appear explicitly.
\begin{align}
\dfrac{D \rho}{D t} + \rho \nabla \cdot \textbf{v} = 0 \label{eq:gov-nc-rho}\\
\dfrac{D \rho \xi}{D t} + \rho \xi \nabla \cdot \textbf{v} = 0 \label{eq:gov-nc-ks}\\
\dfrac{D \textbf{v}}{D t} + \dfrac{\nabla P}{\rho} =\textbf{g} \label{eq:gov-nc-v}\\
\dfrac{D e}{D t} + \dfrac{P \nabla \cdot \textbf{v}}{\rho} = 0 \label{eq:gov-nc-e}
\end{align}
%Governing equations in Lagrangian form, Eq. (\ref{eq:gov-nc-rho}) to Eq.(\ref{eq:gov-nc-e}), and Eulerian form, Eq. (\ref{eq:gov-cs-rho}) to Eq. (\ref{eq:gov-cs-e}), are actually equivalent, as deriving the Lagrangian form from the Eulerian is purely based on mathematics without any additional physical assumptions.

@@

\section{SPH method} \label{sec:SPH_method}
SPH is a mesh-free Lagragian method. In SPH, the domain (control mass) is discretized by a set of particles or discretization points and the position of each particle (discretization point) is updated at every time step based on the motion computed. Approximation of all field variables (velocity, density and pressure, ect.) is obtained by interpolation based on discretization points. The physical laws (such as conservation laws of mass, momentum and energy) written in the form of PDEs (partial differential equations)  or ODEs (ordinary differential equations) need to be transformed into the Lagrangian particle formalism of SPH. Using a kernel function that provides the weighted estimation of the field variables at any point, the integral equations are estimated as sums over particles in a compact subdomain defined by the support of the kernel function associated with the discretization points. Thus, field variables associated to the particle are updated based on its neighbors. Each kernel function has a compact support determined by smoothing length of each particle. There are several review papers by \citet{monaghan1992smoothed, monaghan2005smoothed, rosswog2009astrophysical, price2012smoothed, monaghan2012smoothed}, giving a pretty comprehensive view over SPH. We only refer here to the representation of the constitutive equations in SPH and put more focus on specific numerical techniques for plume modeling.

\subsection{Fundamental principles}
There are several procedures for discretizing governing equations (PDEs or ODEs) with SPH. We present here one of them following \citet{monaghan1992smoothed, monaghan2005smoothed, monaghan2012smoothed}. The starting point of approximating a function with SPH is the translation property of the Dirac function.
\begin{equation}
A\left(\textbf{x}\right)=\int_{-\infty}^{\infty} A\left(\textbf{x} \prime\right) \delta \left(\textbf{x} \prime - \textbf{x}\right) d\textbf{x} \prime
\label{eq:Dirac-translation}
\end{equation}
The Dirac function in Eq. (\ref{eq:Dirac-translation}) can be approximated by a weighting function $w\left(\textbf{x}-\textbf{x}\prime, h\right)$ (or $w\left(\textbf{x}\prime-\textbf{x}, h\right)$)) which tends to a Dirac function when the smoothing length $h \rightarrow 0$ :
\begin{equation}
\lim _{h \rightarrow 0} w\left(\textbf{x} \prime-\textbf{x}, h\right) =  \delta \left(\textbf{x} \prime - \textbf{x}\right)
\label{eq:SPH_kernel_delta}
\end{equation}
An arbitrary function $A\left(\textbf{x}\right)$ can then be approximated by:
\begin{equation}
A\left(\textbf{x}\right) \approx <A\left(\textbf{x}\right)> = \int_{\Omega} A\left(\textbf{x} \prime\right) w\left(\textbf{x}-\textbf{x}\prime, h\right) d\textbf{x}\prime + O\left(h^2\right)
\label{eq:SPH-fundamental-principle}
\end{equation}
Where $h$ is the smoothing length, determining the interaction distance. As the weighting function is symmetric (Eq. (\ref{eq:SPH-kernel-symmetric})) and satisfies the normalization condition (Eq. (\ref{eq:SPH-kernel-normalization-prop})), odd error terms in Eq. (\ref{eq:SPH-fundamental-principle}) vanishe leading to a second order approximation. However, in practice, second order of accuracy can not be achieved because there is no guarantee on the symmetry of particle distribution in real simulation \citep{price2012smoothed}.
Recall that $d\textbf{x}\prime = \dfrac{dm \prime}{\rho \prime}$, the integration equation, Eq. (\ref{eq:SPH-fundamental-principle}), can be approximated by summation and lead to an approximation of the function $A$:
\begin{equation}
<A\left(\textbf{x}\right)> \approx \sum_b m_b \dfrac{A_b}{\rho_b} w\left(\textbf{x}-\textbf{x}_b, h\right)
\label{eq:SPH-approximation-sum}
\end{equation}
where the summation is over all the particles within the region of compact support of the weighting function. 
Gradient terms may be straightforwardly calculated by taking the derivative of Eq. (\ref{eq:SPH-approximation-sum}), giving
\begin{equation}
\begin{split}
<\nabla A\left(\textbf{x}\right)> & = \dfrac{\partial }{\partial \textbf{x}} \int_{\Omega} A\left(\textbf{x} \prime\right) w\left(\textbf{x}-\textbf{x}\prime, h\right) d\textbf{x}\prime + O\left(h^2\right) \\
& \approx \sum_b m_b \dfrac{A_b}{\rho_b} \nabla w\left(\textbf{x} - \textbf{x}_b, h\right)
\end{split} 
\label{eq:SPH-scalar-function-gradient}
\end{equation}
For vector quantities the expressions are similar, simply replacing $A$ with $\textbf{A}$ in Eq. (\ref{eq:SPH-approximation-sum}) and Eq. (\ref{eq:SPH-scalar-function-gradient}), giving
\begin{align}
<\textbf{A}\left(\textbf{x}\right)> \approx \sum_b m_b \dfrac{\textbf{A}_b}{\rho_b} w\left(\textbf{x}-\textbf{x}_b, h\right) \\
<\nabla \cdot \textbf{A}\left(\textbf{x}\right)> \approx \sum_b m_b \dfrac{\textbf{A}_b}{\rho_b} \cdot \nabla w\left(\textbf{x} - \textbf{x}_b, h\right) \\
<\nabla \times \textbf{A}\left(\textbf{x}\right)> \approx \sum_b m_b \dfrac{\textbf{A}_b}{\rho_b} \times \nabla w\left(\textbf{x} - \textbf{x}_b, h\right) \\
<\nabla^j \textbf{A}^i\left(\textbf{x}\right)> \approx \sum_b m_b \dfrac{\textbf{A}_b^i}{\rho_b} \nabla^j w\left(\textbf{x} - \textbf{x}_b, h\right) 
\label{eq:SPH-vecctor-function}
\end{align}
%
